---
title: "quality control and pre-processing"
author: "Chiara_Caprioli"
date: "29/12/2022"
output:
  html_document:
    code_folding: hide
---
## Set up
```{r, setup, include=FALSE}
# set working directory
knitr::opts_knit$set(root.dir = '/Users/ieo4874/Desktop/working/SCMseq/')
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages, echo=F, message=F, warning=F}
#load packages
library(tidyverse)
library(Seurat)
library(scran)
library(scater)
library(patchwork)
library(viridis)
library(ggforce)
library(gghalves)
library(ggridges)
library(scDblFinder)
library(SeuratObject)
```

```{r}
# set paths
path_main <- '/Users/ieo4874/Desktop/working/SCMseq/'
path_data <- paste0(path_main, 'data/')
path_results <- paste0(path_main, 'results_and_plots/qc_pp/')
```

```{r}
# samples
samples <- c('AML4', 'AML5', 'sAML1', 'AML2', 'AML3', 'sAML4')
all <- c(samples, 'healthy BM')
```

```{r, warning=FALSE, message=FALSE}
#set colors 
colors <- list()

colors$sample <- setNames(
  c("#227093", "#ffb142", "#485460", '#F9791EFF', '#C0D8D8FF', '#009E73FF', '#F0F0F0'), 
  all
)

colors$SRSF2_state <- setNames(
  c("#FF9932FF", "#405D61FF", '#F0F0F0'),
  c("SRSF2 mutated", "SRSF2 unmutated", 'healthy BM')
)

colors$cell_cycle <- setNames(
  c("#45aaf2", "#f1c40f", "#e74c3c"),
  c("G1", "S", "G2M")
)

colors$discrete <- c(
  '#046C9AFF', '#D69C4EFF', '#899DA4FF', '#C93312FF', '#D8B70AFF',
  '#02401BFF', '#A2A475FF', '#81A88DFF', '#972D15FF', '#CADDE1FF',
  '#7B906FFF', '#174D79FF',  '#898989FF', '#D4CDB1FF', '#708090FF',
  '#6CA9C3FF', '#3A3533FF', '#000E33FF', '#800000FF',  '#175E78FF')
```

## Helper functions
```{r, warning=FALSE, message=FALSE}
niceFormat <- function(number) {
  formatC(number, format = "f", big.mark = ",", digits = 0)
}

load10xData <- function(
  sample_name,
  path,
  max_number_of_cells = NULL
) {

  ## read transcript count matrix
  transcript_counts <- list.files(
      path,
      recursive = TRUE,
      full.names = TRUE
    ) %>%
    .[grep(., pattern = 'mtx')] %>%
    Matrix::readMM()

  ## read cell barcodes and assign as column names of transcript count matrix
  colnames(transcript_counts) <- list.files(
      path,
      recursive = TRUE,
      full.names = TRUE
    ) %>%
    .[grep(., pattern = 'barcodes')] %>%
    .[grep(., pattern = 'tsv')] %>%
    read_tsv(col_names = FALSE, col_types = cols()) %>%
    pull(1) %>%
    gsub(., pattern = '-[0-9]{1,2}', replacement = '') %>%
    paste0(., '-', sample_name)

  ## read gene names
  gene_names <- list.files(
      path,
      recursive = TRUE,
      full.names = TRUE
    ) %>%
    .[grep(., pattern = 'genes|features')] %>%
    read_tsv(col_names = FALSE, col_types = cols()) %>%
    pull(ifelse(ncol(.) > 1, 2, 1))

  ## if necessary, merge counts from different transcripts of the same gene
  if ( any(duplicated(gene_names)) == TRUE ) {

    ## get names of genes with multiple entries
    duplicated_gene_names <- unique(gene_names[which(duplicated(gene_names))])

    ## log message
    message(
      glue::glue(
        "{format(Sys.time(), '[%Y-%m-%d %H:%M:%S]')} Summing up counts for ",
        "{length(duplicated_gene_names)} genes with multiple entries."
      )
    )

    ## go through genes with multiple entries
    for ( i in duplicated_gene_names ) {

      ## extract transcripts counts for current gene
      tmp_counts <- transcript_counts[which(gene_names == i),]

      ## make sure at least 2 rows are present
      if ( nrow(tmp_counts) >= 2 ) {

        ## sum up counts in each cell
        tmp_counts_sum <- Matrix::colSums(tmp_counts)

        ## remove original counts from transcript matrix and the gene name from
        ## the list of gene names
        transcript_counts <- transcript_counts[-c(which(gene_names == i)),]
        gene_names <- gene_names[-c(which(gene_names == i))]

        ## add summed counts to end of transcript count matrix and current gene
        ## name to end of gene names
        transcript_counts <- rbind(transcript_counts, tmp_counts_sum)
        gene_names <- c(gene_names, i)
      }
    }
  }

  ## assign gene names as row names
  rownames(transcript_counts) <- gene_names

  ## down-sample cells if more cells than `max_number_of_cells` are present in
  ## count matrix
  if (
    !is.null(max_number_of_cells) &&
    is.numeric(max_number_of_cells) &&
    max_number_of_cells < ncol(transcript_counts)
  ) {
    temp_cells_to_keep <- base::sample(1:ncol(transcript_counts), max_number_of_cells)
    transcript_counts <- transcript_counts[,temp_cells_to_keep]
  }

  ##
  message(
    glue::glue(
      "{format(Sys.time(), '[%Y-%m-%d %H:%M:%S]')} Loaded counts for sample ",
      "{sample_name} with {niceFormat(ncol(transcript_counts))} cells and ",
      "{niceFormat(nrow(transcript_counts))} genes."
    )
  )

  ##
  return(transcript_counts)
}

sameGeneNames <- function(counts) {

  ## create place holder for gene names from each count matrix
  gene_names <- list()

  ## go through count matrices
  for ( i in names(counts) ) {

    ## extract gene names
    gene_names[[i]] <- rownames(counts[[i]])
  }

  ## check if all gene names are the same (using the first matrix as a reference
  ## for the others)
  return(all(sapply(gene_names, FUN = identical, gene_names[[1]])))
}

averagenCountnFeature <- function(cells) {
  output <- tibble(
    'sample' = character(),
    'mean(nCount)' = numeric(),
    'median(nCount)' = numeric(),
    'mean(nFeature)' = numeric(),
    'median(nFeature)' = numeric()
  )
  for ( i in levels(cells$sample) ) {
    tmp <- tibble(
      'sample' = i,
      'mean(nCount)' = cells %>% filter(sample == i) %>% pull(nCount) %>% mean(),
      'median(nCount)' = cells %>% filter(sample == i) %>% pull(nCount) %>% median(),
      'mean(nFeature)' = cells %>% filter(sample == i) %>% pull(nFeature) %>% mean(),
      'median(nFeature)' = cells %>% filter(sample == i) %>% pull(nFeature) %>% median()
    )
    output <- bind_rows(output, tmp)
  }
  return(output)
}
```

## Load samples
```{r}
sample_names <- list.dirs(paste0(path_main, 'matrices'), recursive = FALSE) %>% basename()

transcripts <- list(
  raw = list()
)

for ( i in sample_names ) {
  transcripts$raw[[i]] <- load10xData(
    i, paste0('matrices/', i, '/'),
    max_number_of_cells = NULL
  )
}
```

## Check gene names
```{r}
sameGeneNames(transcripts$raw) #FALSE
```

```{r}
add_1_names <- rownames(transcripts$raw$AML2)[which(!rownames(transcripts$raw$AML2) %in% rownames(transcripts$raw$AML4))] #4683
add_2_names <- rownames(transcripts$raw$AML4)[which(!rownames(transcripts$raw$AML4) %in% rownames(transcripts$raw$AML2))] #1606

add_1 <- matrix(data = NA, 
                nrow = length(add_1_names), 
                ncol = ncol(transcripts$raw$AML4), dimnames = NULL)
rownames(add_1) <- add_1_names
transcripts$raw$AML4 <- rbind(transcripts$raw$AML4, add_1)
dim(transcripts$raw$AML4)

add_1 <- matrix(data = NA, 
                nrow = length(add_1_names), 
                ncol = ncol(transcripts$raw$AML5), dimnames = NULL)
rownames(add_1) <- add_1_names
transcripts$raw$AML5 <- rbind(transcripts$raw$AML5, add_1)
dim(transcripts$raw$AML5)

add_1 <- matrix(data = NA, 
                nrow = length(add_1_names), 
                ncol = ncol(transcripts$raw$sAML1), dimnames = NULL)
rownames(add_1) <- add_1_names
transcripts$raw$sAML1 <- rbind(transcripts$raw$sAML1, add_1)
dim(transcripts$raw$sAML1)

add_2 <- matrix(data = NA, 
                nrow = length(add_2_names), 
                ncol = ncol(transcripts$raw$AML2), dimnames = NULL)
rownames(add_2) <- add_2_names
transcripts$raw$AML2 <- rbind(transcripts$raw$AML2, add_2)
dim(transcripts$raw$AML2)

add_2 <- matrix(data = NA, 
                nrow = length(add_2_names), 
                ncol = ncol(transcripts$raw$AML3), dimnames = NULL)
rownames(add_2) <- add_2_names
transcripts$raw$AML3 <- rbind(transcripts$raw$AML3, add_2)
dim(transcripts$raw$AML3)

add_2 <- matrix(data = NA, 
                nrow = length(add_2_names), 
                ncol = ncol(transcripts$raw$sAML4), dimnames = NULL)
rownames(add_2) <- add_2_names
transcripts$raw$sAML4 <- rbind(transcripts$raw$sAML4, add_2)
dim(transcripts$raw$sAML4)

transcripts$raw$AML4 <- transcripts$raw$AML4[match(rownames(transcripts$raw$AML2), rownames(transcripts$raw$AML4)),]
transcripts$raw$AML5 <- transcripts$raw$AML5[match(rownames(transcripts$raw$AML2), rownames(transcripts$raw$AML5)),]
transcripts$raw$sAML1 <- transcripts$raw$sAML1[match(rownames(transcripts$raw$AML2), rownames(transcripts$raw$sAML1)),]
```

```{r}
sameGeneNames(transcripts$raw) #TRUE
```

```{r}
transcripts$raw$merged <- do.call(cbind, transcripts$raw)
transcripts$raw$merged[ is.na(transcripts$raw$merged) ] <- 0
dim(transcripts$raw$merged)
```

## Quality control
```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}
cells <- tibble(
  cell = colnames(transcripts$raw$merged),
  sample = colnames(transcripts$raw$merged) %>%
    strsplit("-") %>%
    vapply(FUN.VALUE = character(1), `[`, 2) %>%
    factor(levels = samples),
  nCount = colSums(transcripts$raw$merged),
  nFeature = colSums(transcripts$raw$merged != 0)
)

DataFrame(cells)
```

### add column with mitochondrial genes (HUMAN)
```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}
mitochondrial_genes_here <- read_tsv(paste0(path_main, "genes_mt_hg_name.tsv.gz"),
  col_names = FALSE
) %>%
  filter(X1 %in% rownames(transcripts$raw$merged)) %>%
  pull(X1)

cells$percent_MT <-
  Matrix::colSums(transcripts$raw$merged[mitochondrial_genes_here,]) /
  Matrix::colSums(transcripts$raw$merged)

DataFrame(cells)
```

### find and remove doublets
```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}
sce <- SingleCellExperiment(assays = list(counts = transcripts$raw$merged))
sce <- scDblFinder(sce)

cells$multiplet_class <- colData(sce)$scDblFinder.class

cells %>%
  filter(multiplet_class != 'singlet') %>%
  group_by(sample) %>%
  summarize(count = n()) %>%
  ggplot(aes(x = sample, y = count, fill = sample)) +
  geom_col(color = 'black') +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells$sample))) +
  scale_y_continuous(name = 'Number of doublets', labels = scales::comma) +
  theme(
    axis.title.y = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = 'none'
  ) +
  coord_flip()

ggsave(paste0(path_results, 'plots/qc_number_of_doublets_by_sample.png'), height = 3, width = 6
)
```

```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}
p1 <- ggplot(cells, aes(x = multiplet_class, y = nCount, fill = multiplet_class)) +
  geom_violin(draw_quantiles = c(0.5), scale = 'area', trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$discrete) +
  scale_x_discrete(limits = rev(levels(cells$multiplet_class))) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = 'Number of transcripts', subtitle = 'linear scale') +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = 'none'
  ) +
  coord_flip()
p2 <- ggplot(cells, aes(x = multiplet_class, y = nCount, fill = multiplet_class)) +
  geom_violin(draw_quantiles = c(0.5), scale = 'area', trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$discrete) +
  scale_x_discrete(limits = rev(levels(cells$multiplet_class))) +
  scale_y_log10(labels = scales::comma) +
  labs(title = 'Number of transcripts', subtitle = 'log-scale') +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = 'none'
  ) +
  coord_flip()
p3 <- ggplot(cells, aes(x = multiplet_class, y = nFeature, fill = multiplet_class)) +
  geom_violin(draw_quantiles = c(0.5), scale = 'area', trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$discrete) +
  scale_x_discrete(limits = rev(levels(cells$multiplet_class))) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = 'Number of expressed genes', subtitle = 'linear scale') +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = 'none'
  ) +
  coord_flip()
p4 <- ggplot(cells, aes(x = multiplet_class, y = nFeature, fill = multiplet_class)) +
  geom_violin(draw_quantiles = c(0.5), scale = 'area', trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$discrete) +
  scale_x_discrete(limits = rev(levels(cells$multiplet_class))) +
  scale_y_log10(labels = scales::comma) +
  labs(title = 'Number of expressed genes', subtitle = 'log-scale') +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = 'none'
  ) +
  coord_flip()

ggsave(paste0(path_results, "plots/qc_nCount_nFeature_by_multiplet_class.png"),
  p1 + p3 + p2 + p4 + plot_layout(ncol = 2),
  height = 7, width = 10
)
```

```{r}
cells <- filter(cells, multiplet_class == "singlet")
```

### find and remove low quality cells

#### check before filtering
```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}

p1 <- ggplot(cells, aes(x = sample, y = nCount, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells$sample))) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Number of transcripts", subtitle = "linear scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()
p2 <- ggplot(cells, aes(x = sample, y = nCount, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells$sample))) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Number of transcripts", subtitle = "log-scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()
p3 <- ggplot(cells, aes(x = sample, y = nFeature, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells$sample))) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Number of expressed genes", subtitle = "linear scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()
p4 <- ggplot(cells, aes(x = sample, y = nFeature, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells$sample))) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Number of expressed genes", subtitle = "log-scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()
p5 <- ggplot(cells, aes(x = sample, y = percent_MT, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells$sample))) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Percent MT transcripts", subtitle = "linear scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()

ggsave(paste0(path_results,"plots/qc_histogram_nCount_nFeature_percentMT_before_filtering_1.png"),
  p1 + p3 + p5 +
    p2 + p4 + plot_layout(ncol = 3),
  height = 7, width = 10)

```

```{r}
p1 <- ggplot(cells, aes(nCount, fill = sample)) +
  geom_histogram(bins = 200) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = 'Number of transcripts', subtitle = 'linear scale', y = 'Number of cells') +
  theme(legend.position = 'none', axis.title.x = element_blank())

p2 <- ggplot(cells, aes(nCount, fill = sample)) +
  geom_histogram(bins = 200) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_log10(labels = scales::comma) +
  labs(title = 'Number of transcripts', subtitle = 'log-scale', y = 'Number of cells') +
  theme(legend.position = 'none', axis.title.x = element_blank())

p3 <- ggplot(cells, aes(nFeature, fill = sample)) +
  geom_histogram(bins = 200) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_continuous(labels = scales::comma) +
  labs(title = 'Number of expressed genes', subtitle = 'linear scale', y = 'Number of cells') +
  theme(legend.position = 'none',axis.title.x = element_blank())

p4 <- ggplot(cells, aes(nFeature, fill = sample)) +
  geom_histogram(bins = 200) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_log10(labels = scales::comma) +
  labs(title = 'Number of expressed genes', subtitle = 'log-scale', y = 'Number of cells', fill = 'Sample') +
  theme(legend.position = 'none', axis.title.x = element_blank())

p5 <- ggplot(cells, aes(percent_MT, fill = sample)) +
  geom_histogram(bins = 200) +
  theme_bw() +
  scale_x_continuous(labels = scales::percent) +
  scale_fill_manual(name = 'Sample', values = colors$sample) +
  labs(title = 'Percent MT transcripts', subtitle = 'linear scale', y = 'Number of cells') +
  theme(legend.position = 'right', axis.title.x = element_blank(), legend.text = element_text(size = 8))

ggsave(paste0(path_results, 'plots/qc_histogram_ncount_nfeature_percentmt_before_filtering_3.png'),
  p1 + p3 + p5 +
  p2 + p4 + plot_layout(ncol = 3),
  height = 7, width = 10
)
```

#### Plot nFeature/nCount
```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}

cells %>% 
  ggplot(aes(nCount, nFeature, color = percent_MT)) +
  geom_point(size = 0.1) +
  scale_x_continuous(name = "Number of transcripts", labels = scales::comma) +
  scale_y_continuous(name = "Number of expressed genes", labels = scales::comma) +
  theme_bw() +
  scale_color_viridis(
    name = "Percent MT\ntranscripts",
    limits = c(0,1),
    labels = scales::percent,
    guide = guide_colorbar(frame.colour = "black", ticks.colour = "black")
  )

ggsave(paste0(path_results, "plots/qc_nFeature_over_nCount_before_filtering_mt.png"), height = 4, width = 6)

cells %>% 
  ggplot(aes(nCount, nFeature, color = sample)) +
  geom_point(size = 0.1) +
  scale_x_continuous(name = "Number of transcripts", labels = scales::comma) +
  scale_y_continuous(name = "Number of expressed genes", labels = scales::comma) +
  theme_bw() +
  scale_color_manual(name = 'Sample', values = colors$sample) +
  guides(colour = guide_legend(override.aes = list(size=3)))

ggsave(paste0(path_results, "plots/qc_nFeature_over_nCount_before_filtering_samples.png"), height = 4, width = 6)
```

#### Define adaptive thresholds for filtering
```{r, include=TRUE, warning=FALSE, message=FALSE}
cells <- cells 

median_nCount <- median(cells$nCount) 

mad_nCount <- mad(cells$nCount) 

median_nFeature <- median(cells$nFeature) 

mad_nFeature <- mad(cells$nFeature) 

median_percent_MT <- median(cells$percent_MT) 

mad_percent_MT <- mad(cells$percent_MT)
```

```{r, include=TRUE, warning=FALSE, message=FALSE}
median_mads <- as.data.frame(list(
  median_nCount = median_nCount,
  mad_nCount = mad_nCount,
  median_nFeature = median_nFeature,
  mad_nFeature = mad_nFeature,
  median_percent_MT = median_percent_MT,
  mad_percent_MT = mad_percent_MT
)) 

write_csv(median_mads, paste0(path_results, 'tables/median_mads_merged.csv'))
```

#### Set thresholds 
It is generally advised to stay within 3-5 MADs around the median.
```{r, warning=FALSE, message=FALSE}

thresholds_nCount <- c(250, median_nCount + 5*mad_nCount)

thresholds_nFeature <- c(500, median_nFeature + 3*mad_nFeature) 

thresholds_percent_MT <- c(0, median_percent_MT + 3*mad_percent_MT)
```

```{r, include=TRUE, warning=FALSE, message=FALSE}
thresholds <- as.data.frame(list(
  thresholds_nCount = thresholds_nCount,
  thresholds_nFeature = thresholds_nFeature,
  thresholds_percent_MT = thresholds_percent_MT
)) 

write_csv(thresholds, paste0(path_results, 'tables/thresholds_merged.csv'))
```

#### Plot thresholds 
```{r}
p1 <- ggplot(cells, aes(nCount, fill = sample)) +
  geom_histogram(binwidth = 20) +
  geom_vline(xintercept = median_nCount, color = "cyan") +
  geom_vline(xintercept = thresholds_nCount, color = "red") +
  theme_bw() +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(name = 'Sample', values = colors$sample) +
  labs(title = "Number of transcripts") +
  facet_wrap(~sample, nrow = length(samples)) +
  theme(legend.position = 'none')

p2 <- ggplot(cells, aes(nFeature, fill = sample)) +
  geom_histogram(binwidth = 20) +
  geom_vline(xintercept = median_nFeature, color = "cyan") +
  geom_vline(xintercept = thresholds_nFeature, color = "red") +
  theme_bw() +
  scale_y_continuous(labels = scales::comma) +
  scale_fill_manual(name = 'Sample', values = colors$sample) +
  labs(title = "Number of expressed genes") +
  facet_wrap(~sample, nrow = length(samples)) +
  theme(legend.position = 'none')

p3 <- ggplot(cells, aes(percent_MT, fill = sample)) +
  geom_histogram(binwidth = 0.005) +
  geom_vline(xintercept = median_percent_MT, color = "cyan") +
  geom_vline(xintercept = thresholds_percent_MT, color = "red") +
  theme_bw() +
  scale_fill_manual(name = 'Sample', values = colors$sample) +
  labs(title = "Percent MT transcripts") +
  facet_wrap(~sample, nrow = length(samples)) +
  theme(legend.position = 'none')

ggsave(paste0(path_results, "plots/qc_histogram_nCount_nFeature_percentMT_thresholds.png"),
  p1 + p2 + p3 + plot_layout(ncol = 3), height = 18, width = 12
)
```

```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}
p1 <- ggplot(cells, aes(x = sample, y = nCount, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  geom_hline(yintercept = median_nCount, color = "black") +
  geom_hline(yintercept = thresholds_nCount, color = "red") +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells$sample))) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Number of transcripts", subtitle = "linear scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()
p2 <- ggplot(cells, aes(x = sample, y = nCount, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  geom_hline(yintercept = median_nCount, color = "black") +
  geom_hline(yintercept = thresholds_nCount, color = "red") +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells$sample))) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Number of transcripts", subtitle = "log-scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()
p3 <- ggplot(cells, aes(x = sample, y = nFeature, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  geom_hline(yintercept = median_nFeature, color = "black") +
  geom_hline(yintercept = thresholds_nFeature, color = "red") +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells$sample))) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Number of expressed genes", subtitle = "linear scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()
p4 <- ggplot(cells, aes(x = sample, y = nFeature, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  geom_hline(yintercept = median_nFeature, color = "black") +
  geom_hline(yintercept = thresholds_nFeature, color = "red") +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells$sample))) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Number of expressed genes", subtitle = "log-scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()
p5 <- ggplot(cells, aes(x = sample, y = percent_MT, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  geom_hline(yintercept = median_percent_MT, color = "black") +
  geom_hline(yintercept = thresholds_percent_MT, color = "red") +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells$sample))) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Percent MT transcripts", subtitle = "linear scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()

ggsave(paste0(path_results, "plots/qc_vln_nCount_nFeature_percentMT_thresholds.png"),
  p1 + p3 + p5 +
    p2 + p4 + plot_layout(ncol = 3),
  height = 7, width = 10
)
```

```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}
ggplot(cells, aes(nCount, nFeature, color = percent_MT)) +
  geom_point(size = 0.5) +
  geom_hline(yintercept = thresholds_nFeature, color = "red") +
  geom_vline(xintercept = thresholds_nCount, color = "red") +
  scale_x_continuous(name = "Number of transcripts", labels = scales::comma) +
  scale_y_continuous(name = "Number of expressed genes", labels = scales::comma) +
  theme_bw() +
  scale_color_viridis(
    name = "Percent MT\ntranscripts",
    limits = c(0,1),
    labels = scales::percent,
    guide = guide_colorbar(frame.colour = "black", ticks.colour = "black")
  ) + 
  facet_wrap(~sample, ncol = length(samples)*0.5, nrow = 2)

ggsave(paste0(path_results, "plots/qc_nFeature_over_nCount_thresholds.png"), height = 6, width = 10)
```

#### Filter cells 
```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}
cells_filtered <- cells %>%
  dplyr::filter(
    nCount >= thresholds_nCount[1],
    nCount <= thresholds_nCount[2],
    nFeature >= thresholds_nFeature[1],
    nFeature <= thresholds_nFeature[2],
    percent_MT >= thresholds_percent_MT[1],
    percent_MT <= thresholds_percent_MT[2]
  )

DataFrame(cells_filtered)
```

#### Check after filtering out low-quality cells
```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}
p1 <- ggplot(cells_filtered, aes(x = sample, y = nCount, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells_filtered$sample))) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Number of transcripts", subtitle = "linear scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()
p2 <- ggplot(cells_filtered, aes(x = sample, y = nCount, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells_filtered$sample))) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Number of transcripts", subtitle = "log-scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()
p3 <- ggplot(cells_filtered, aes(x = sample, y = nFeature, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells_filtered$sample))) +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Number of expressed genes", subtitle = "linear scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()
p4 <- ggplot(cells_filtered, aes(x = sample, y = nFeature, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells_filtered$sample))) +
  scale_y_log10(labels = scales::comma) +
  labs(title = "Number of expressed genes", subtitle = "log-scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()
p5 <- ggplot(cells_filtered, aes(x = sample, y = percent_MT, fill = sample)) +
  geom_violin(draw_quantiles = c(0.5), scale = "area", trim = FALSE) +
  theme_bw() +
  scale_fill_manual(values = colors$sample) +
  scale_x_discrete(limits = rev(levels(cells_filtered$sample))) +
  scale_y_continuous(labels = scales::percent) +
  labs(title = "Percent MT transcripts", subtitle = "linear scale") +
  theme(
    axis.title = element_blank(),
    panel.grid.major.y = element_blank(),
    panel.grid.minor.y = element_blank(),
    legend.position = "none"
  ) +
  coord_flip()

ggsave(paste0(path_results, "plots/qc_histogram_nCount_nFeature_percentMT_filtered.png"),
  p1 + p3 + p5 +
    p2 + p4 + plot_layout(ncol = 3),
  height = 7, width = 10
)
```

```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}
ggplot(cells_filtered, aes(nCount, nFeature, color = percent_MT)) +
  geom_point(size = 0.5) +
  scale_x_continuous(name = "Number of transcripts", labels = scales::comma) +
  scale_y_continuous(name = "Number of expressed genes", labels = scales::comma) +
  theme_bw() +
  scale_color_viridis(
    name = "Percent MT\ntranscripts",
    limits = c(0,1),
    labels = scales::percent,
    guide = guide_colorbar(frame.colour = "black", ticks.colour = "black")
  ) +
  facet_wrap(~sample, ncol = length(samples)*0.5, nrow = 2)

ggsave(paste0(path_results, "plots/qc_nFeature_over_nCount_filtered.png"),  height = 6, width = 10)
```

```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}
cells_to_keep <- cells_filtered$cell

length(cells_to_keep) # 21566
```

### remove genes which are expressed in fewer than 5 cells
maybe better sample-wise?
```{r, warning=FALSE, message=FALSE}
genes <- tibble(
  gene = rownames(transcripts$raw$merged),
  count = Matrix::rowSums(transcripts$raw$merged),
  cells = Matrix::rowSums(transcripts$raw$merged != 0)
)
genes_to_keep <- genes %>% dplyr::filter(cells >= 5) %>% pull(gene)
length(genes_to_keep) # 25026
```

## Generate filtered high-quality matrix
```{r, warning=FALSE, message=FALSE}
transcripts$raw$filtered <- transcripts$raw$merged[genes_to_keep,cells_to_keep]
```

```{r, warning=FALSE, message=FALSE}
cells_qc <- tibble(
  sample = character(),
  n_cells_before = numeric(),
  n_cells_after = numeric()
)
for ( i in samples ) {
  tmp <- tibble(
    sample = i,
    n_cells_before = grep(colnames(transcripts$raw$merged), pattern = paste0("-", i, "$"))
    %>% length(),
    n_cells_after = grep(colnames(transcripts$raw$filtered), pattern = paste0("-", i,
                                                                      "$")) %>% length()
  )
  cells_qc <- bind_rows(cells_qc, tmp)
}

write_csv(cells_qc, paste0(path_results, 'tables/cells_before_after_qc.csv'))
```

```{r}
saveRDS(transcripts$raw$filtered, paste0(path_data, 'merged_matrix_filtered.rds'))
```

## Create Seurat object of AML samples (merged)
```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}
merged_AML <- CreateSeuratObject(
  counts = transcripts$raw$filtered,
  min.cells = 0,
  min.features = 0
)

#store sample assignment for every cell in the meta data
merged_AML@meta.data$sample <- Cells(merged_AML) %>%
  strsplit("-") %>%
  vapply(FUN.VALUE = character(1), `[`, 2) %>%
  factor(levels = samples)
```

## QC healthy BM from Human Cell Atlas
```{r}
#healthy BM: single merged Seurat object with 8 donors (not normalized)
library(SeuratData)
InstallData('hcabm40k')
data("hcabm40k")
library(hcabm40k.SeuratData)
```

```{r}
unique(Idents(hcabm40k))
hcabm40k <- subset(x = hcabm40k, downsample = 3000)
hcabm40k
```

```{r}
hcabm40k@meta.data$sample <- hcabm40k@meta.data$orig.ident
hcabm40k@meta.data$orig.ident <- 'healthy BM'

DefaultAssay(object = hcabm40k) <- "RNA"
hcabm40k[["percent_MT"]] <- PercentageFeatureSet(hcabm40k, "MT-")
```

```{r}
median_nCount <- median(hcabm40k$nCount_RNA) 
mad_nCount <- mad(hcabm40k$nCount_RNA) 

median_nFeature <- median(hcabm40k$nFeature_RNA) 
mad_nFeature <- mad(hcabm40k$nFeature_RNA) 

median_percent_MT <- median(hcabm40k$percent_MT) 
mad_percent_MT <- mad(hcabm40k$percent_MT)

median_mads <- as.data.frame(list(
  median_nCount = median_nCount,
  mad_nCount = mad_nCount,
  median_nFeature = median_nFeature,
  mad_nFeature = mad_nFeature,
  median_percent_MT = median_percent_MT,
  mad_percent_MT = mad_percent_MT
)) 

write_csv(median_mads, paste0(path_results, 'tables/median_mads_hca_BM.csv'))
```

```{r}
summary(hcabm40k$nFeature_RNA) 
```

```{r, warning=FALSE, message=FALSE}
thresholds_nCount <- c(250, 50000)

thresholds_nFeature <- c(500, 6000) 

thresholds_percent_MT <- c(0, 25)
```

```{r, include=TRUE, warning=FALSE, message=FALSE}
thresholds <- as.data.frame(list(
  thresholds_nCount = thresholds_nCount,
  thresholds_nFeature = thresholds_nFeature,
  thresholds_percent_MT = thresholds_percent_MT
)) 

write_csv(thresholds, paste0(path_results, 'tables/thresholds_hca_BM.csv'))
```

```{r}
p1 <- ggplot(hcabm40k@meta.data, aes(nCount_RNA)) +
  geom_histogram(binwidth = 20) +
  geom_vline(xintercept = median_nCount, color = "cyan") +
  geom_vline(xintercept = thresholds_nCount, color = "red") +
  theme_bw() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Number of transcripts") +
  theme(legend.position = 'none')

p2 <- ggplot(hcabm40k@meta.data, aes(nFeature_RNA)) +
  geom_histogram(binwidth = 20) +
  geom_vline(xintercept = median_nFeature, color = "cyan") +
  geom_vline(xintercept = thresholds_nFeature, color = "red") +
  theme_bw() +
  scale_y_continuous(labels = scales::comma) +
  labs(title = "Number of expressed genes") +
  theme(legend.position = 'none')

p3 <- ggplot(hcabm40k@meta.data, aes(percent_MT)) +
  geom_histogram(binwidth = 0.005) +
  geom_vline(xintercept = median_percent_MT, color = "cyan") +
  geom_vline(xintercept = thresholds_percent_MT, color = "red") +
  theme_bw() +
  labs(title = "Percent MT transcripts") +
  theme(legend.position = 'none')

ggsave(paste0(path_results, "plots/qc_histogram_nCount_nFeature_percentMT_thresholds_hca_BM.png"),
  p1 + p2 + p3 + plot_layout(ncol = 3), height = 3, width = 12
)
```

```{r}
cells_filtered <- hcabm40k@meta.data %>%
  dplyr::filter(
    nCount_RNA >= thresholds_nCount[1],
    nCount_RNA <= thresholds_nCount[2],
    nFeature_RNA >= thresholds_nFeature[1],
    nFeature_RNA <= thresholds_nFeature[2],
    percent_MT >= thresholds_percent_MT[1],
    percent_MT <= thresholds_percent_MT[2]
  )

cells_filtered$percent_MT <- cells_filtered$percent_MT*0.01

DataFrame(cells_filtered)
```

```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}
ggplot(cells_filtered, aes(nCount_RNA, nFeature_RNA, color = percent_MT)) +
  geom_point(size = 0.5) +
  scale_x_continuous(name = "Number of transcripts", labels = scales::comma) +
  scale_y_continuous(name = "Number of expressed genes", labels = scales::comma) +
  theme_bw() +
  scale_color_viridis(
    name = "Percent MT\ntranscripts",
    limits = c(0,1),
    labels = scales::percent,
    guide = guide_colorbar(frame.colour = "black", ticks.colour = "black")
  ) 

ggsave(paste0(path_results, "plots/qc_nFeature_over_nCount_filtered_hca_BM.png"),  height = 6, width = 10)
```

```{r}
hcabm40k <- hcabm40k[,rownames(cells_filtered)]
```

## Merge AML samples with healthy BM
```{r}
merged_AML@meta.data$orig.ident <- 'AML'
hcabm40k@meta.data$sample <- 'healthy BM'

merged_seurat <- merge(merged_AML, y = hcabm40k, project = "SCMseq")
merged_seurat$sample <- factor(merged_seurat$sample, levels = all)

table(merged_seurat$orig.ident)
```

```{r, warning=FALSE, message=FALSE, out.width="75%", fig.align="center"}
#check again the distribution of number of transcripts and expressed genes per cell by sample 
temp_labels <- merged_seurat@meta.data %>%
  group_by(sample) %>%
  tally()

p1 <- ggplot() +
  geom_half_violin(
    data = merged_seurat@meta.data, aes(sample, nCount_RNA, fill = sample),
    side = "l", show.legend = FALSE, trim = FALSE
  ) +
  geom_half_boxplot(
    data = merged_seurat@meta.data, aes(sample, nCount_RNA, fill = sample),
    side = "r", outlier.color = NA, width = 0.4, show.legend = FALSE
  ) +
  geom_text(
    data = temp_labels,
    aes(x = sample, y = -Inf, label = paste0("n = ", format(n, big.mark = ",",
                                                            trim = TRUE)), vjust = -1),
    color = "black", size = 2.8
  ) +
  scale_color_manual(values = colors$sample) +
  scale_fill_manual(values = colors$sample) +
  scale_y_continuous(labels = scales::comma, expand = c(0.08,0)) +
  theme_bw() +
  labs(x = "", y = "Number of transcripts") +
  theme(
    panel.grid.major.x = element_blank(),
    axis.title.x = element_blank()
  )

p2 <- ggplot() +
  geom_half_violin(
    data = merged_seurat@meta.data, aes(sample, nFeature_RNA, fill = sample),
    side = "l", show.legend = FALSE, trim = FALSE
  ) +
  geom_half_boxplot(
    data = merged_seurat@meta.data, aes(sample, nFeature_RNA, fill = sample),
    side = "r", outlier.color = NA, width = 0.4, show.legend = FALSE
  ) +
  geom_text(
    data = temp_labels,
    aes(x = sample, y = -Inf, label = paste0("n = ", format(n, big.mark = ",",
                                                            trim = TRUE)), vjust = -1),
    color = "black", size = 2.8
  ) +
  scale_color_manual(values = colors$sample) +
  scale_fill_manual(values = colors$sample) +
  scale_y_continuous(
    name = "Number of expressed genes",
    labels = scales::comma, expand = c(0.08,0)
  ) +
  theme_bw() +
  theme(
    panel.grid.major.x = element_blank(),
    axis.title.x = element_blank()
  )

ggsave(paste0(path_results, "plots/nCount_nFeature_by_sample.png"),
  p1 + p2 + plot_layout(ncol = 2), height = 4, width = 10
)
```

## Add percentage of mitochondrial genes in metadata
```{r}
DefaultAssay(object = merged_seurat) <- "RNA"
merged_seurat[["percent_MT"]] <- PercentageFeatureSet(merged_seurat, "MT-")
```

```{r}
saveRDS(merged_seurat, paste0(path_data, 'merged_seurat_counts.rds'))
```

## Log normalization (for cell lineage mapping)
```{r}
merged_seurat <- readRDS("~/Desktop/working/SCMseq/data/qc_pp/merged_seurat_counts.rds")

merged_seurat
```

```{r}
merged_seurat <- NormalizeData(object = merged_seurat, 
                               normalization.method = "LogNormalize",
                               scale.factor = 10000)
merged_seurat <- FindVariableFeatures(object = merged_seurat, selection.method = "vst")
merged_seurat <- ScaleData(object = merged_seurat,
                           features = NULL, #variable features by default
                           vars.to.regress = 'percent_MT')
```

```{r}
saveRDS(merged_seurat, paste0(path_data, 'merged_seurat_logcounts.rds'))
```

```{r}
sessionInfo()
```
